---
output: 
  github_document:
    df_print: tibble
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  cache = FALSE
)
library(neonstore)
Sys.setenv("NEONSTORE_HOME" = tempfile())
Sys.setenv("NEONSTORE_DB" = tempfile())

```

# neonstore

<!-- badges: start -->
[![R build status](https://github.com/cboettig/neonstore/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/neonstore/actions)
[![Codecov test coverage](https://codecov.io/gh/cboettig/neonstore/branch/master/graph/badge.svg)](https://app.codecov.io/gh/cboettig/neonstore?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/neonstore)](https://CRAN.R-project.org/package=neonstore)
<!-- badges: end -->

`neonstore` provides quick access and persistent storage of NEON data tables. 
`neonstore` emphasizes simplicity and a clean data provenance trail, see 
Provenance section below.

## Installation

Install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/neonstore")
```
## Quickstart

```{r}
library(neonstore)
```

Discover data products of interest:

```{r}
products <- neon_products()

i <- grepl("Populations", products$themes)
products[i, c("productCode", "productName")]
 
i <- grepl("bird", products$keywords)
products[i, c("productCode", "productName")]
```

Download all data files in the bird survey data products.  

```{r}
neon_download("DP1.10003.001")

```


View your store of NEON products:

```{r }
neon_index()
```

These files will persist between sessions, so you only need to download once
or to retrieve updates. `neon_index()` can take arguments to filter by product
or pattern (regular expression) in table name, e.g. `neon_index(table = "brd")`.

Once you determine the table of interest, you can
read in all the component tables into a single `data.frame`

```{r}
neon_read("brd_countdata-expanded")
```



## Database backend

`neonstore` now supports a backend relation database as well. Import data from
the raw downloaded files using `neon_store()`:

```{r}
neon_store(table = "brd_countdata-expanded")
```

Alternately, we could import all data tables associated with a given product:

```{r}
neon_store(product = "DP1.10003.001")
```

Access an imported table using `neon_table()` instead of `neon_read()`:

```{r}
neon_table("brd_countdata")
```




Access the remote database using `neon_db()`.  This is a `DBIConnection` that
can easily be used with `dplyr` functions like `tbl()` or `filter()`.  
Remember that `dplyr` translates these into SQL queries that run directly
on the database.  

```{r}
library(dplyr)
con <- neon_db()
brd <- tbl(con, "brd_countdata-basic-DP1.10003.001")
brd %>% filter(siteID == "ORNL")

```

Note that we need to include the product name in the table name when accessing the database, as table names alone may not be unique. RStudio users can also list and explore all tables interactively in the Connections pane in RStudio using `neon_pane()`.  


## Note on API limits

If `neon_download()` exceeds the API request limit (with or without the token),
`neonstore` will simply pause for the required amount of time to avoid 
rate-limit-based errors.  

[The NEON API now rate-limits requests.](https://data.neonscience.org/data-api/rate-limiting/#api-tokens).
Using a personal token will increase the number of requests you can make before
encountering this delay.  See link for directions on registering for a token.
Then pass this token in `.token` argument of `neon_download()`,
or for frequent use, add this token as  an environmental variable, `NEON_DATA`
to your local `.Renviron` file in your user's home directory. 
`neon_download()` must first query each the API of each NEON site which collects
that product, for each month the product is collected.


(It would be much more efficient on the NEON server if the API could take
queries of the from `/data/<product>/<site>`, and pool the results, rather than
require each month of sampling separately!)

```{r include=FALSE}
unlink("my_neondata.zip")
unlink("index.csv")
Sys.unsetenv("NEONSTORE_HOME")
Sys.unsetenv("NEONSTORE_DB")
```

## Non-stacking files and low-level interface

At it's core, `neonstore` is simply a mechanism to download files from the NEON API.
While the `.csv` files from the Observation Systems (OS, e.g. bird count surveys),
and Instrument Systems (e.g. aquatic sensors) are typically stacked into large
tables, other products, such as the `.laz` and `.tif` images produced by the
airborne observation platform LIDAR and cameras may require a different approach.



```{r}
# Read in a large file list for illustration purposes
cper_data <- readr::read_csv("https://minio.thelio.carlboettiger.info/shared-data/neon_data_catalog.csv.gz")

## Typically one would read all files in local store, e.g. list.file(neon_dir())
df <- neon_filename_parser(cper_data$name)
```

```{r}
library(dplyr)
df %>% count(EXT, sort=TRUE)
```

We can take a look at all `laz` LIDAR files:

```{r}
df %>% 
  filter(EXT == "laz")
```

Note that many of the airborne observation platform (AOP) products, such as
these LIDAR files, do not include the PRNUM or REV components that make up part
of the `productCode`s used in the NEON `product` tables.  


