---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# neonstore

<!-- badges: start -->
<!-- badges: end -->

The goal of neonstore is to provide access and persistent storage of NEON data tables

## Installation


Install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/neonstore")
```
## Example

Download all CSV files in the bird survey data products.  This will omit downloads of any existing data files. 

```{r example, cache=TRUE}
library(neonstore)
neon_download("DP1.10003.001", file_regex = "[.]csv")

```


Now, view your library of NEON products.  These will persist between sessions, 
so you only need to download once, or to retrieve updates.

```{r }
neon_tables()
```
See files associated with a given table, 

```{r output="hide"}
meta <- neon_index("brd_countdata") 
meta
```

```{r echo=FALSE}
# pretty-print
tibble::as_tibble(meta)
```

Read in all the component tables into a single data.frame

```{r}
brd_countdata <- neon_read(meta)
```

